---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file. -->
<!-- Re-render with devtools::build_readme(). -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  # 8in, not the 7in default: the dumbbell subtitle names the CI level and
  # the chance line, and clips against the legend at 7in.
  fig.width = 8,
  fig.height = 4.5,
  dpi = 150
)
# The bootstrap is random; seed it so re-rendering the README produces the
# same numbers and doesn't churn the diff.
set.seed(2026)
```

# cindexdecomp

> C-Index Decomposition for Survival Models

<!-- badges: start -->
[![R-CMD-check](https://github.com/Lainsm/cindexdecomp/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/Lainsm/cindexdecomp/actions/workflows/R-CMD-check.yaml)
[![pkgdown](https://github.com/Lainsm/cindexdecomp/actions/workflows/pkgdown.yaml/badge.svg)](https://github.com/Lainsm/cindexdecomp/actions/workflows/pkgdown.yaml)
[![Codecov test coverage](https://codecov.io/gh/Lainsm/cindexdecomp/graph/badge.svg)](https://app.codecov.io/gh/Lainsm/cindexdecomp)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/licenses/MIT)
<!-- badges: end -->

## Overview

Harrell's concordance index pools two different ranking tasks:

- **Event-Event pairs (`C_ee`)** — both subjects had the event. The harder task.
- **Event-Censored pairs (`C_ec`)** — one had the event, one was censored later.

As censoring rises the event-event pairs vanish and `C_ec` dominates the
pooled score, so a model close to chance on true events can still report a
stable global C-index. `cindexdecomp` makes that visible.

## Installation

```r
# install.packages("pak")
pak::pak("Lainsm/cindexdecomp")
```

## Quick start

```{r quickstart, message = FALSE}
library(cindexdecomp)
library(survival)

lung2 <- lung[complete.cases(lung), ]
cox <- coxph(Surv(time, status - 1) ~ age + sex + ph.ecog, data = lung2)

fit <- decompose_cindex(
  lung2$time,
  lung2$status - 1, # lung codes 1/2; the package expects 0/1
  predict(cox, type = "lp"),
  n_boot = 1000
)
fit
```

```{r quickstart-confint}
confint(fit)
```

A formula interface is available too, and takes the data frame directly:

```{r formula}
lung2$lp <- predict(cox, type = "lp")
decompose_cindex(Surv(time, status - 1) ~ lp, data = lung2)
```

## What the numbers say

On this cohort the global C-index looks respectable, but the split shows
where it comes from: the event-censored pairs score materially higher than
the event-event pairs, and the masking gap quantifies the difference.

```{r gap}
c(C_ee = fit$C_ee, C_ec = fit$C_ec, C_global = fit$C_global, gap = fit$gap)
```

Because `confint()` bootstraps the gap itself — resampling subjects, not
pairs — you can say whether that gap is distinguishable from zero.

## The identity

Any concordance index that averages over comparable pairs,

```
C_w = sum(w_ij * c_ij) / sum(w_ij)
```

decomposes exactly:

```
C_w = (W_ee * C_ee + W_ec * C_ec) / (W_ee + W_ec)
```

where `W` denotes sums of weights. Setting `w_ij = 1` recovers Harrell's C.
The identity is verifiable from the returned object:

```{r identity}
with(fit, (W_ee * C_ee + W_ec * C_ec) / (W_ee + W_ec)) - fit$C_global
```

## Weightings

| Constructor | Estimator |
|---|---|
| `weights_harrell()` | Harrell's C |
| `weights_uno()` | Uno's IPCW C |
| `weights_truncated(tau)` | C truncated at `tau` |
| `weights_custom(fn, name)` | any user-supplied rule |

```{r weightings}
decompose_cindex(
  lung2$time, lung2$status - 1, lung2$lp,
  weights = weights_uno()
)
```

The pooled result agrees with `survival::concordance()` to machine
precision for Harrell's weighting under every tie pattern, and for Uno's
weighting when event times are distinct. Under tied times, Uno's
pairwise estimator diverges from the counting-process form in
`survival` by
around 1e-04 — immaterial in practice, and explained in the vignette
(`vignette("cindexdecomp")`).

## Tracing the masking

`censoring_curve()` re-applies the decomposition under a sweep of
administrative cut-offs, so you can watch `C_global` hold steady while
`C_ee` moves:

```{r curve, fig.alt = "Global C-index and event-event C-index plotted against the cohort censoring rate. The global curve stays roughly flat near 0.65 while the event-event curve falls towards and below 0.5, and the shaded band between them is the masking gap."}
curve <- censoring_curve(lung2$time, lung2$status - 1, lung2$lp)
autoplot(curve)
```

The global curve holds near 0.65 across the sweep while the event-event
curve falls to chance and below. That divergence is the masking the
package is named for. `autoplot()` is re-exported, so **ggplot2 does not
need to be attached** to get the plot — only to modify it afterwards.

## Comparing models

```{r compare, fig.alt = "Dumbbell plot with one row per model, each row spanning that model's event-event and event-censored C-index with bootstrap error bars."}
risks <- list(
  age_only = predict(coxph(Surv(time, status - 1) ~ age, data = lung2)),
  full = lung2$lp
)

cmp <- compare_decompositions(risks, lung2$time, lung2$status - 1, n_boot = 200)
cmp
autoplot(cmp)
```

## Functions

| Function | Description |
|---|---|
| `decompose_cindex()` | Decomposes into `C_ee` and `C_ec` |
| `confint()` | Bootstrap intervals, including for the masking gap |
| `censoring_curve()` | Traces the decomposition as censoring rises |
| `compare_decompositions()` | Runs several models over one cohort |
| `autoplot()` | Plots any of the above |
| `theme_cindex()` | Light (default) and dark plot themes |

## Model agnostic

Any model that produces a risk score works:

```r
decompose_cindex(time, status, predict(cox_fit, type = "lp"))
decompose_cindex(time, status, predict(xgb_fit, xgb.DMatrix(X)))
decompose_cindex(time, status, predict(rsf_fit, newdata = test)$predicted)
```

## Limitations

- Assumes non-informative censoring.
- `C_ee` is imprecise when few event-event pairs exist. Use `confint()`;
  pair counts describe composition, not precision.
- Evaluates discrimination only, not calibration.
- Model-based estimators such as Gönen–Heller are out of scope: they
  integrate over an assumed model rather than counting observed pairs, so
  the event-event partition is undefined for them.

## Citation

```{r citation, comment = ""}
citation("cindexdecomp")
```

## References

Harrell FE et al. (1982) Evaluating the yield of medical tests.
*JAMA*, 247(18), 2543-2546.

Uno H et al. (2011) On the C-statistics for evaluating overall adequacy of
risk prediction procedures with censored survival data.
*Statistics in Medicine*, 30(10), 1105-1117.
