---
title: "cindexdecomp — full API walkthrough on survival::lung"
format:
  html:
    embed-resources: true
    toc: true
    code-fold: false
execute:
  warning: true
---

Exploratory notebook exercising every exported function of `cindexdecomp`
against `survival::lung`. Not a package vignette — lives in `dev/`, ignored
by `R CMD build` (see `.Rbuildignore`).

```{r setup}
devtools::load_all(here::here())
library(survival)
set.seed(1)
```

## The data

`lung$status` is coded 1 = censored, 2 = dead; `cindexdecomp` wants
0/1 with 1 = event.

```{r data}
lung2 <- lung
lung2$event <- lung2$status - 1L
# age, ph.ecog have missing values; keep only complete cases for the
# risk scores used below.
lung2 <- lung2[stats::complete.cases(lung2[, c("time", "event", "age", "ph.ecog", "sex")]), ]
nrow(lung2)
table(lung2$event)
```

Two risk scores to compare: age alone, and age + ECOG performance status
(both "higher is riskier" on the natural scale, no need to negate).

```{r risk-scores}
risk_age   <- lung2$age
risk_full  <- lung2$age + 10 * lung2$ph.ecog
```

## `weights_*()` constructors

```{r weights}
weights_harrell()
weights_uno()
weights_uno(tau = 300)
weights_truncated(tau = 300)
weights_custom(function(t, G) rep(1, length(t)), name = "UnitCustom")
```

## `decompose_cindex()` — default (vector) method

```{r decompose-default}
fit_harrell <- decompose_cindex(lung2$time, lung2$event, risk_full, n_boot = 300)
fit_harrell
```

```{r decompose-uno}
fit_uno <- decompose_cindex(
  lung2$time, lung2$event, risk_full,
  weights = weights_uno(), n_boot = 300
)
fit_uno
```

## `decompose_cindex()` — formula method

```{r decompose-formula}
fit_formula <- decompose_cindex(
  Surv(time, event) ~ risk_full,
  data = data.frame(lung2, risk_full = risk_full),
  n_boot = 100
)
fit_formula
```

## `print()`, `summary()`, `as.data.frame()` on a `cindex_decomp`

```{r methods-decomp}
print(fit_harrell)
summary(fit_harrell)
as.data.frame(fit_harrell)
```

## `confint.cindex_decomp()`

```{r confint}
confint(fit_harrell)
confint(fit_harrell, level = 0.80)
confint(fit_harrell, parm = "gap")
```

## `censoring_curve()`

```{r censoring-curve}
curve <- censoring_curve(lung2$time, lung2$event, risk_full, n_thresholds = 8)
curve
as.data.frame(curve)
```

## `compare_decompositions()`

Two risk scores on the same cohort, Harrell weighting.

```{r compare}
cmp <- compare_decompositions(
  list(age_only = risk_age, age_ecog = risk_full),
  lung2$time, lung2$event,
  n_boot = 300
)
cmp
as.data.frame(cmp)
```

A matrix of risk scores also works (Finding 8 of the final review).

```{r compare-matrix}
risk_mat <- cbind(age_only = risk_age, age_ecog = risk_full)
compare_decompositions(risk_mat, lung2$time, lung2$event, n_boot = 0)
```

## `autoplot()` — all three plot methods

```{r autoplot-decomp}
autoplot(fit_harrell)
```

```{r autoplot-curve}
autoplot(curve)
```

```{r autoplot-compare}
autoplot(cmp)
```

## `theme_cindex()`

```{r theme}
autoplot(fit_harrell) + theme_cindex()
```

## Error handling / validation surface

Each of these is expected to error, not silently misbehave — confirms the
validation added across the redesign and the final review's fix wave.

```{r errors}
try(decompose_cindex(time = lung2$time, status = lung2$event, risk = risk_full))
try(decompose_cindex(lung2$time, lung2$event, risk_full, higher_is_riskier = 0))
try(confint(fit_harrell, level = -0.2))
try(censoring_curve(lung2$time, lung2$event, risk_full, higher_is_riskier = "yes"))
try(compare_decompositions(matrix(1:10, ncol = 2), lung2$time[1:5], lung2$event[1:5]))
try(decompose_cindex(c(lung2$time, NA), c(lung2$event, 1), c(risk_full, 1)))
```

## Session info

```{r session}
sessionInfo()
```
