---
title: "Decomposing the C-index"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Decomposing the C-index}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  # 8in, not the 7in default: the dumbbell subtitle names the CI level and
  # the chance line, and clips against the legend at 7in.
  fig.width = 8, fig.height = 4.5
)
```

```{r setup}
library(cindexdecomp)
library(survival)
```

## The problem

Harrell's concordance index pools two different ranking tasks. Some
comparable pairs consist of two subjects who both had the event; others
pair an event with a subject censored later. The second task is easier,
and as censoring rises it comes to dominate the pooled score: a model can
report a stable, respectable global C-index while doing no better than
chance at ordering the events it was built to predict.

## The decomposition

Write any concordance index as a weighted average over comparable pairs,

$$C_w = \frac{\sum_{ij} w_{ij} c_{ij}}{\sum_{ij} w_{ij}},$$

where $c_{ij} \in \{0, 0.5, 1\}$ records whether the pair was ranked
correctly. Splitting the pairs by the status of the later subject gives

$$C_w = \frac{W_{ee} C_{ee} + W_{ec} C_{ec}}{W_{ee} + W_{ec}},$$

with $W$ denoting sums of weights. The identity holds exactly for every
weighting -- it is an algebraic regrouping of the same sum, not an
approximation.

```{r}
set.seed(7)
n <- 400
risk <- rnorm(n)
event_time <- rexp(n, rate = exp(0.8 * risk) * 0.1)
cens_time <- rexp(n, rate = 0.08)
time <- pmin(event_time, cens_time)
status <- as.numeric(event_time <= cens_time)

fit <- decompose_cindex(time, status, risk, n_boot = 200)
fit
```

The global figure agrees with `survival::concordance()` exactly:

```{r}
c(
  cindexdecomp = fit$C_global,
  survival = concordance(Surv(time, status) ~ risk, reverse = TRUE)$concordance
)
```

## Ties

Three cases arise, and the package follows `survival::concordance()` in
each:

* two events at the same recorded time are **not** comparable;
* an event and a censoring at the same recorded time **are** comparable,
  with the event treated as occurring first;
* equal risk scores receive half credit.

The second case matters in practice because survival times are usually
recorded in whole days, so exact ties are common rather than exceptional.

## Choosing a weighting

The weighting selects the estimator. One engine produces all of them.

```{r}
weightings <- list(
  Harrell   = weights_harrell(),
  Uno       = weights_uno(),
  Truncated = weights_truncated(tau = quantile(time[status == 1], 0.5))
)

do.call(rbind, lapply(names(weightings), function(nm) {
  f <- decompose_cindex(time, status, risk, weights = weightings[[nm]])
  data.frame(
    weighting = nm, C_global = f$C_global,
    C_ee = f$C_ee, C_ec = f$C_ec
  )
}))
```

`C_global` varies with the estimator chosen, and every one of those values
is defensible. `C_ee` does not: it stays close to the same figure
throughout, well below `C_ec` in every row. Uno's inverse-probability
weighting corrects the bias in the pooled estimate; it does not reveal the
composition problem underneath. The masking is a property of the data, not
of the estimator.

A user-supplied weighting needs no changes to the package:

```{r}
own <- weights_custom(function(t, G) 1 / G(t), name = "IPCW (power 1)")
decompose_cindex(time, status, risk, weights = own)$C_global
```

### Uno's C under tied event times

`cindexdecomp` implements Uno's estimator in its published **pairwise**
form: every comparable pair is weighted by $1/\hat{G}(T_i)^2$, where
$\hat G$ is the Kaplan-Meier estimate of the censoring distribution.
`survival::concordance(timewt = "n/G2")` uses a counting-process
formulation that applies its weight per event *time* instead. The two
agree to the limit of double-precision arithmetic when event times are
distinct, and diverge materially once times are tied:

```{r}
tie_agreement <- function(time, status, risk) {
  ref_h <- concordance(Surv(time, status) ~ risk, reverse = TRUE)$concordance
  ref_u <- concordance(Surv(time, status) ~ risk,
    reverse = TRUE,
    timewt = "n/G2"
  )$concordance
  fit_h <- decompose_cindex(time, status, risk, weights = weights_harrell())
  fit_u <- decompose_cindex(time, status, risk, weights = weights_uno())
  c(harrell = fit_h$C_global - ref_h, uno = fit_u$C_global - ref_u)
}

# Day-scale rounding mimics how survival time is actually recorded (whole
# days), which is what makes exact ties common rather than exceptional.
time_days <- round(time * 30) + 1
lung2 <- lung[complete.cases(lung), ]

tie_raw <- rbind(
  `Continuous times`   = tie_agreement(time, status, risk),
  `Day-scale, rounded` = tie_agreement(time_days, status, risk),
  `survival::lung`     = tie_agreement(lung2$time, lung2$status - 1, lung2$age)
)

tie_tab <- data.frame(
  Data = rownames(tie_raw),
  `Harrell agreement` = formatC(tie_raw[, "harrell"], format = "e", digits = 2),
  `Uno agreement` = formatC(tie_raw[, "uno"], format = "e", digits = 2),
  check.names = FALSE, row.names = NULL
)
knitr::kable(tie_tab, align = c("l", "r", "r"))
```

The largest deviation observed here (`r formatC(max(abs(tie_raw[, "uno"])), format = "e", digits = 2)`,
in the `survival::lung` row) is roughly two orders of magnitude below the
sampling variability of a global estimate on data of this size -- the
bootstrap standard deviation of `C_global` for the fit above is
`r formatC(sd(fit$boot[, "C_global"]), format = "e", digits = 2)`. It does
not affect any practical conclusion. It is documented because a user
cross-checking against `survival` on tied data would otherwise reasonably
suspect a bug.

The Harrell column is zero to floating-point precision throughout: exactly
`0` for continuous times and for `survival::lung`, and a single unit of
rounding error (`r formatC(tie_raw["Day-scale, rounded", "harrell"], format = "e", digits = 2)`,
i.e. one bit) once times are tied and rounded to whole days. That is
agreement to the limit of double-precision arithmetic, not an
approximation.

Per-pair weighting is a requirement here rather than a preference: a
weight attached to an event *time* cannot be partitioned into
event-event and event-censored pair sets, so the decomposition this
package computes could not be defined under the counting-process
formulation. Harrell's weighting is unaffected and matches
`survival::concordance()` exactly under every tie pattern.

## Inference

Intervals come from resampling **subjects**, not pairs. `r n` subjects
generate `r format(fit$N_ee + fit$N_ec, big.mark = ",")` comparable pairs,
but the information content is `r n` observations: resampling pairs
directly gives intervals several times too narrow.

```{r}
confint(fit)
```

The final row is usually the one of interest. It is computed inside each
replicate, so the correlation between `C_ee` and `C_ec` is carried through
rather than assumed away.

## Censoring curves

`censoring_curve()` applies a series of administrative cut-offs to the
**whole cohort** and recomputes at each. The reported rate is therefore the
cohort's actual censoring rate, including subjects censored before any
cut-off was applied.

```{r}
cv <- censoring_curve(time, status, risk, n_thresholds = 12)
head(as.data.frame(cv))
```

```{r}
autoplot(cv)
```

Estimates resting on few event-event pairs are flagged in
`low_precision` and drawn faintly. They are never removed, and no value is
discarded for being small: `r sum(as.data.frame(cv)$low_precision)` of the
`r nrow(as.data.frame(cv))` thresholds above are flagged, all at the
high-censoring end of the sweep where the flag is meant to bite.

## Comparing models

```{r}
risks <- list(
  full = risk,
  noisy = risk + rnorm(n, sd = 1.5)
)
cmp <- compare_decompositions(risks, time, status, n_boot = 200)
cmp
```

```{r}
autoplot(cmp)
```

Adding noise to the risk score shrinks `C_ee`, `C_ec` and the gap between
them together -- a reminder that the gap is a property of a specific
model's failure mode on this cohort, not a universal constant to be
compared across unrelated studies.

## A worked example

```{r}
cox <- coxph(Surv(time, status - 1) ~ age + sex + ph.ecog, data = lung2)
decompose_cindex(
  Surv(time, status - 1) ~ predict(cox, type = "lp"),
  data = lung2,
  n_boot = 200
)
```

Note the `status - 1`: `lung` codes status as
`r paste(sort(unique(lung$status)), collapse = "/")`, while the package
expects 0/1. Data already coded 0/1 needs no adjustment.

## Scope

The decomposition applies to any estimator that averages over observed
comparable pairs. Model-based estimators such as Gönen–Heller integrate
over an assumed model rather than counting pairs, so the event-event and
event-censored partition is undefined for them; they are outside the scope
of this package by construction.

Non-informative censoring is assumed throughout. The package evaluates
discrimination only, not calibration.
