---
title: "Check your C-index"
subtitle: "The pooled concordance index hides its own composition"
author: 
  - name: "Malik J. Lainsbury"
    affiliation: "University of Cambridge"
date: 2026-09-19
format:
  revealjs:
    theme: [default, symposium.scss]
    embed-resources: true
    slide-number: c/t
    # Canonical CRAN reference form. NOTE: dead until CRAN accepts the
    # package; swap for the live pkgdown vignette if presenting before then:
    #   lainsm.github.io/cindexdecomp/articles/cindexdecomp.html
    footer: "CRAN.R-project.org/package=cindexdecomp"
    # The `#` section breaks nest their slides as vertical stacks; linear mode
    # makes left/right step through every slide in order, so a right-arrow
    # during the talk cannot skip a section's contents.
    navigation-mode: linear
    transition: none
    fig-align: center
    # auto-stretch rewrites a slide's lone image as .r-stretch and sizes it to
    # the leftover vertical space. On these slides that computed to height 0
    # and the figures vanished. The figures are deliberately sized below, so
    # the feature is off and CSS caps them instead.
    auto-stretch: false
    # Revealjs numbers code lines and gives every block a scrollbar by
    # default; on a one-line snippet both are noise, and the scrollbar renders
    # bright against the dark ground.
    code-line-numbers: false
    code-overflow: wrap
    fig-width: 10
    fig-height: 4.6
    fig-dpi: 200
    df-print: kable
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
    dev: ragg_png
bibliography: /home/malik/Documents/uni/cindexdecomp/cindexdecomp.bib
---

```{r}
#| label: setup
#| include: false

library(survival)
library(ggplot2)
pkgload::load_all(here::here(), quiet = TRUE)

# Mirrors cindexdecomp:::cindex_palette(dark = TRUE); duplicated rather than
# reached for with ::: so the deck does not depend on a non-exported function.
# The two pair-type colours are identical in both variants -- only the
# neutrals change -- so the deck's palette still matches package output.
pal <- list(
  ee     = "#DC267F",
  ec     = "#648FFF",
  global = "#F8F8F2",
  fg     = "#F8F8F2",
  bg     = "#282A36",
  grid   = "#3A3C4E",
  muted  = "#9AA0B8"
)

theme_talk <- function(base_size = 18) {
  theme_cindex(dark = TRUE, base_size = base_size) +
    theme(
      plot.title.position = "plot",
      legend.position = "top",
      legend.title = element_blank(),
      legend.text = element_text(size = rel(0.95))
    )
}
```

```{r}
#| label: cohorts
#| include: false
#| cache: true

# Five published cohorts, all shipped with `survival`, ordered by censoring
# rate. Cached because five bootstraps at n_boot = 1000 is minutes of compute
# and this deck gets re-rendered constantly during rehearsal.
set.seed(2026)
N_BOOT <- 1000

decomp_of <- function(label, disease, time, status, lp) {
  f <- decompose_cindex(time, status, lp, n_boot = N_BOOT)
  ci <- confint(f)
  data.frame(
    cohort    = label,
    disease   = disease,
    n         = f$n,
    cens      = 100 * (1 - f$n_events / f$n),
    ee_share  = 100 * f$W_ee / (f$W_ee + f$W_ec),
    C_ee      = f$C_ee,
    C_ec      = f$C_ec,
    C_global  = f$C_global,
    gap       = f$gap,
    gap_lo    = ci["gap", 1],
    gap_hi    = ci["gap", 2]
  )
}

lung2 <- lung[complete.cases(lung), ]
m_lung <- coxph(Surv(time, status - 1) ~ age + sex + ph.ecog, data = lung2)

colon2 <- colon[
  colon$etype == 2 &
    complete.cases(colon[, c(
      "age", "sex", "obstruct", "nodes", "differ", "extent", "rx"
    )]),
]
m_colon <- coxph(
  Surv(time, status) ~ age + sex + obstruct + nodes + differ + extent + rx,
  data = colon2
)

m_gbsg <- coxph(
  Surv(rfstime, status) ~
    age + meno + size + grade + nodes + pgr + er + hormon,
  data = gbsg
)

m_rott <- coxph(
  Surv(dtime, death) ~
    age + meno + size + grade + nodes + pgr + er + hormon + chemo,
  data = rotterdam
)

flc <- flchain[
  flchain$futime > 0 &
    complete.cases(flchain[, c("age", "sex", "kappa", "lambda", "mgus")]),
]
m_flc <- coxph(
  Surv(futime, death) ~ age + sex + kappa + lambda + mgus,
  data = flc
)

ladder <- rbind(
  decomp_of(
    "lung", "NSCLC", lung2$time, lung2$status - 1,
    predict(m_lung, type = "lp")
  ),
  decomp_of(
    "colon", "Colon cancer", colon2$time, colon2$status,
    predict(m_colon, type = "lp")
  ),
  decomp_of(
    "gbsg", "Breast cancer", gbsg$rfstime, gbsg$status,
    predict(m_gbsg, type = "lp")
  ),
  decomp_of(
    "rotterdam", "Breast cancer", rotterdam$dtime, rotterdam$death,
    predict(m_rott, type = "lp")
  ),
  decomp_of(
    "flchain", "General population", flc$futime, flc$death,
    predict(m_flc, type = "lp")
  )
)
ladder <- ladder[order(ladder$cens), ]
rownames(ladder) <- NULL

fmt <- function(x, d = 3) formatC(x, format = "f", digits = d)
pc <- function(x, d = 1) paste0(formatC(x, format = "f", digits = d), "%")
```

```{r}
#| label: cohort-years
#| include: false

# Deliberately NOT inside the cached `cohorts` chunk: adding metadata there
# would change that chunk's code and invalidate five bootstraps of compute.
# rotterdam and flchain carry their own study years; the other three are from
# the survival package's documentation (Loprinzi 1994; Laurie 1989; gbsg is
# documented as a 1984-1989 trial).
yrs <- c(
  lung      = "1994",
  colon     = "1989",
  gbsg      = "1984-1989",
  rotterdam = paste(range(rotterdam$year), collapse = "-"),
  flchain   = paste(range(flc$sample.yr), collapse = "-")
)
ladder$years <- unname(yrs[ladder$cohort])
stopifnot(!anyNA(ladder$years))
```

```{r}
#| label: fns-tasks
#| include: false

# Slide 5's timeline figure, built in stages so the versions can be stacked as
# reveal.js fragments and appear to draw themselves. Layout is identical at
# every stage -- later elements are drawn transparent rather than omitted --
# so nothing shifts as the build advances.
patients <- data.frame(
  id    = factor(c("B", "A", "D", "C"), levels = c("B", "A", "D", "C")),
  pair  = c("Pair 1", "Pair 1", "Pair 2", "Pair 2"),
  stop  = c(400, 200, 800, 200),
  event = c(TRUE, TRUE, FALSE, TRUE)
)
patients$panel <- factor(
  ifelse(
    patients$pair == "Pair 1",
    "Pair 1  ·  both died",
    "Pair 2  ·  one died, one left alive"
  ),
  levels = c("Pair 1  ·  both died", "Pair 2  ·  one died, one left alive")
)

task_plot <- function(stage = 3) {
  d <- patients
  d$vis <- ifelse(d$pair == "Pair 1" | stage >= 2, 1, 0)
  d$col <- ifelse(d$pair == "Pair 1", pal$ee, pal$ec)

  # y = 1.5 sits midway between the two patients in each panel. Each panel
  # carries exactly two discrete levels under scales = "free_y", so anything
  # above 2 would be clipped.
  labs_df <- data.frame(
    panel = factor(levels(patients$panel), levels = levels(patients$panel)),
    x = 880,
    y = 1.5,
    lab = c("ORDERING\nhard", "DETECTION\neasy"),
    col = c(pal$ee, pal$ec),
    vis = if (stage >= 3) 1 else 0
  )

  ggplot(d, aes(y = id)) +
    geom_segment(
      aes(x = 0, xend = stop, yend = id, colour = col, alpha = vis),
      linewidth = 2.4, lineend = "round"
    ) +
    geom_point(
      data = d[d$event, ],
      aes(x = stop, colour = col, alpha = vis),
      shape = 4, size = 7, stroke = 2.4
    ) +
    geom_point(
      data = d[!d$event, ],
      aes(x = stop, colour = col, alpha = vis),
      shape = 21, size = 6, stroke = 2.4, fill = pal$bg
    ) +
    geom_text(
      data = labs_df,
      aes(x = x, y = y, label = lab, colour = col, alpha = vis),
      hjust = 1, vjust = 0.5, size = 5.4, fontface = "bold", lineheight = 0.95
    ) +
    scale_colour_identity() +
    scale_alpha_identity() +
    # Panel 2's strip title is blanked until its data appears, so stage 1 does
    # not telegraph the reveal. The strip keeps its height either way, so the
    # layout still does not shift between fragments.
    facet_wrap(
      ~panel, ncol = 1, scales = "free_y",
      labeller = as_labeller(stats::setNames(
        c(levels(patients$panel)[1], if (stage >= 2) levels(patients$panel)[2] else " "),
        levels(patients$panel)
      ))
    ) +
    scale_x_continuous(
      "days since enrolment",
      limits = c(0, 900), breaks = seq(0, 800, 200), expand = expansion(c(0.01, 0.02))
    ) +
    labs(y = NULL) +
    theme_talk(16) +
    theme(
      strip.background = element_blank(),
      # theme_cindex() does not set strip.text, so it keeps theme_classic()'s
      # near-black grey10 and disappears against the dark ground. Set here
      # rather than inherited.
      strip.text = element_text(
        hjust = 0, face = "bold", size = rel(1.0), colour = pal$fg
      ),
      panel.grid.major.y = element_blank(),
      axis.line.y = element_blank(),
      axis.ticks.y = element_blank()
    )
}
```

## The number we all report {.smaller}

85% of ML studies that report a C-index report it as a single number [@odonnellSystematicReviewMachine2025]

::: {.big-stat}
C = 0.65
:::

Read as: *the probability the model ranks a random pair correctly.*

However there is a movement that the c-index might not be the best metric to use in survival analysis [@alabdallahConcordanceIndexDecomposition2024; @blancheCindexNotProper2019;@sierraCindexMultiverse2025]

::: {.fragment .highlight-q}
**Which pairs?**
:::

::: {.notes}
Plant the question here and do not answer it. Everything for the next twelve
minutes is an answer to "which pairs".
:::

## Discrimination is one axis among several {.smaller}

| | what it asks |
|---|---|
| **Brier score / IBS** | are the predicted probabilities accurate? |
| **Time-dependent AUC** | does it discriminate *at this horizon*? |
| **Calibration slope / CITL** | do predicted risks match observed ones? |
| **Royston–Sauerbrei D** | how far apart are the risk groups? |
| **C-index** | does it get the *ordering* right? |

::: {.fragment}
The C-index dominates reporting practice, often as the only number quoted.
:::

::: {.fragment .highlight-q}
This talk is not "C is the wrong metric."<br>
It is: **the pooled form of C hides its own composition.**
:::

::: {.notes}
Scope-setting. Costs 40 seconds and keeps Q&A on topic -- without it someone
spends your question time asking why you ignored calibration.
:::

## Censoring is a partial outcome {.smaller}

A patient who leaves the study alive at day 800 has not gone missing. You
know something real: **they survived at least 800 days.**

::: {.fragment}
To score a model you compare *pairs* of patients. But some pairs cannot be
ordered at all:
:::

::: {.fragment}
> Two patients both left alive: one at 200 days, one at 400.
> Which had the higher risk? **Unanswerable.**
:::

::: {.fragment .highlight-q}
Those pairs are dropped before any score is computed.<br>
The pairs that *survive* that filter are the ones that make your number.
:::

## Two pairs. Two very different questions. {.smaller}

```{r}
#| label: tasks-1
#| fig-height: 4.4
task_plot(1)
```

::: {.notes}
Ask the room out loud: which of these two had the higher risk? Let the silence
sit -- it is genuinely hard. Then advance.
:::

## Two pairs. Two very different questions. {.smaller visibility="uncounted"}

```{r}
#| label: tasks-2
#| fig-height: 4.4
task_plot(2)
```

::: {.notes}
Now ask again for pair 2. The room answers instantly. That gap in response
time IS the talk.
:::

## Two pairs. Two very different questions. {.smaller visibility="uncounted"}

```{r}
#| label: tasks-3
#| fig-height: 4.4
task_plot(3)
```

::: {.fragment .highlight-q}
Different questions, and C averages them into one score.
:::

## The mix is set by your study, not your model {.smaller}

::: {.columns}
::: {.column width="50%"}
**Event-event pairs**

Need *two* events.

As censoring rises they thin out **quadratically**.

::: {.ee-tag}
the hard task
:::
:::

::: {.column width="50%"}
**Event-censored pairs**

Need *one* event.

They persist.

::: {.ec-tag}
the easy task
:::
:::
:::

::: {.fragment .highlight-q}
So the easy task's weight in your average grows with your **censoring rate**:
a property of follow-up and study design, **not of the model you built.**
:::

::: {.notes}
This is the mechanism, complete, with no algebra shown. If they only remember
one slide, it should be the previous one; if they remember two, this one.
:::

# Five cohorts {.section-break}

## Five published cohorts, five Cox models {.smaller}

::: {.centre-table}
```{r}
#| label: cohort-table
tab <- data.frame(
  Cohort = ladder$cohort,
  Setting = ladder$disease,
  Years = ladder$years,
  n = format(ladder$n, big.mark = ","),
  Censoring = pc(ladder$cens)
)
knitr::kable(tab, align = "lllrr")
```
:::

::: {.fragment .highlight-q}
Unrelated diseases. Different covariates. Different decades.
:::

::: {.notes}
The censoring ordering is deliberate but goes unsaid on the slide: mention
verbally that the rows are sorted by censoring rate and that this is the one
thing varying systematically across them.
:::

## How Harrell's C is computed {.smaller}

::: {.steps}
::: {.step}
**1. Form every comparable pair.**
Two patients are comparable when you can say which failed first. Both
censored, or the censored one lost to follow-up too early: not comparable,
and dropped.
:::

::: {.step}
**2. Score each pair.**
`1` if the model gives the higher risk to the one who failed first, `0.5` if
their risks tie, `0` otherwise.
:::

::: {.step}
**3. Average over the pairs.**
Every comparable pair carries **equal weight**.
:::
:::

::: {.fragment .highlight-q}
That equal weight is the whole story: `w = 1` for a pair of deaths and `w = 1`
for a death against a survivor, however differently hard those two
comparisons are.
:::

::: {.notes}
Keep this to about 45 seconds. It exists so that C_global on the next slide
is not a black box, and so the w = 1 in the identity later lands as a choice
rather than as notation.
:::

## Read the global column first {.smaller}

:::{.centre-table}
```{r}
#| label: ladder-global
tab <- data.frame(
  Cohort = ladder$cohort,
  Censoring = pc(ladder$cens),
  `C_global` = fmt(ladder$C_global),
  check.names = FALSE
)
knitr::kable(tab, align = "lll")
```
:::

::: {.fragment .highlight-q .center}
`r fmt(min(ladder$C_global))` to `r fmt(max(ladder$C_global))` as censoring
goes from `r pc(min(ladder$cens), 0)` to `r pc(max(ladder$cens), 0)`.
:::

::: {.notes}
Let them read it as model quality.  Read
alone, this column says the models are improving. Do not correct it yet. The next slide is
the whole talk.
:::




## The pooled score tracks the censoring rate {.smaller}

::: {.centre-table}
```{r}
#| label: global-trend
#| fig-height: 5.3
ggplot(ladder, aes(cens, C_global)) +
  geom_hline(yintercept = 0.5, linetype = "dashed",
             colour = pal$muted, linewidth = 0.5) +
  annotate("text", x = 25, y = 0.477, label = "chance", hjust = 1,
           size = 4.2, colour = pal$muted) +
  # A faint connector in censoring order, not a fitted line. C_global is
  # strictly monotone across all five, so this traces the ordering that is
  # actually there rather than asserting a model.
  geom_line(colour = pal$global, linewidth = 0.6, alpha = 0.35,
            linetype = "22") +
  geom_point(colour = pal$global, size = 6) +
  ggrepel::geom_text_repel(
    aes(label = cohort), size = 4.3, colour = pal$fg, seed = 7,
    nudge_y = -0.05, direction = "x", segment.colour = NA, box.padding = 0.25
  ) +
  scale_x_continuous(
    "Cohort Censoring Rate", labels = function(x) paste0(x, "%"),
    limits = c(20, 80), breaks = seq(20, 80, 10)
  ) +
  scale_y_continuous("Global C-index", limits = c(0.44, 0.85),
                     breaks = seq(0.5, 0.8, 0.1)) +
  theme_talk(17)
```
:::



::: {.notes}
Do not concede anything yet. Let the room accept the rising line as model
quality; the next two slides take it away.
:::

## Now add the event-event column {.smaller visibility="uncounted"}

```{r}
#| label: ladder-full
tab <- data.frame(
  Cohort = ladder$cohort,
  Censoring = pc(ladder$cens),
  `EE share` = pc(ladder$ee_share),
  `C_ee` = fmt(ladder$C_ee),
  `C_ec` = fmt(ladder$C_ec),
  `C_global` = fmt(ladder$C_global),
  `Gap [95% CI]` = paste0(
    fmt(ladder$gap), "  [", fmt(ladder$gap_lo), ", ", fmt(ladder$gap_hi), "]"
  ),
  check.names = FALSE
)
knitr::kable(tab, align = "lrrrrrr")
```

::: {.fragment .highlight-q}
`C_ee` is **flat**: `r fmt(min(ladder$C_ee))` to `r fmt(max(ladder$C_ee))` across
all five. Every one of these models is about equally poor at ordering real
events.
:::

::: {.notes}
Only C_global is strictly monotone. Do NOT claim a monotone trend in the gap:
colon (0.093) and gbsg (0.100) both sit below lung (0.120). If asked, say so
plainly -- the flat C_ee column is the load-bearing fact, not the gap.
:::


## The two halves, side by side {.smaller}

::: {.centre-table}
```{r}
#| label: ee-ec-bars
#| fig-height: 4.3
bars <- rbind(
  data.frame(cohort = ladder$cohort, cens = ladder$cens,
             C = ladder$C_ee, which = "C_ee   event-event pairs"),
  data.frame(cohort = ladder$cohort, cens = ladder$cens,
             C = ladder$C_ec, which = "C_ec   event-censored pairs")
)
# cohorts stay in censoring order; C_ee first so magenta leads each group
bars$cohort <- factor(bars$cohort, levels = ladder$cohort)
bars$which <- factor(bars$which, levels = c(
  "C_ee   event-event pairs", "C_ec   event-censored pairs"
))
lab <- setNames(paste0(ladder$cohort, "\n", pc(ladder$cens, 0)), ladder$cohort)

ggplot(bars, aes(cohort, C, fill = which)) +
  geom_col(position = position_dodge(width = 0.72), width = 0.62) +
  geom_text(
    aes(label = fmt(C, 2)), position = position_dodge(width = 0.72),
    vjust = -0.55, size = 4.1, colour = pal$fg
  ) +
  geom_hline(yintercept = 0.5, linetype = "dashed",
             colour = pal$muted, linewidth = 0.5) +
  scale_fill_manual(values = c(pal$ee, pal$ec)) +
  scale_x_discrete(NULL, labels = lab) +
  # "chance" rides a right-hand axis rather than an in-panel annotation: bars
  # start at zero, so every in-panel position collides with one.
  scale_y_continuous(
    "C-index", limits = c(0, 0.95), breaks = seq(0, 0.8, 0.2),
    expand = expansion(c(0, 0.02)),
    sec.axis = dup_axis(breaks = 0.5, labels = "chance", name = NULL)
  ) +
  theme_talk(16)
```
:::

::: {.fragment .highlight-q}
The blue bars carry the score. The magenta bars barely move, and the
shortest of them belongs to the cohort with the **highest** reported C.
:::

::: {.notes}
Bars start at zero deliberately. If asked why they are not truncated at 0.5:
because a truncated baseline would exaggerate exactly the effect being
claimed, and the effect does not need the help.
:::

```{r}
#| label: punchline-values
#| include: false
f <- ladder[ladder$cohort == "flchain", ]
```

## The same model, scored two ways {.smaller}

[C_global]{.stat-name}

::: {.big-stat}
`r fmt(f$C_global)`
:::

A number anyone would write up as a good model.

::: {.fragment}
[C_ee]{.stat-name .ee-colour}

::: {.big-stat .ee-colour}
`r fmt(f$C_ee)`
:::
On the pairs where both patients actually had the event.
:::

::: {.fragment .highlight-q}
**The honest caveat:** five different diseases, five different covariate
sets. Censoring is confounded with everything.<br>
**The answer:** `C_ee` sits at `r fmt(min(ladder$C_ee))` to `r fmt(max(ladder$C_ee))`
in *all five*. No disease-specific story accounts for that.
:::

::: {.notes}
Concede the confounding BEFORE they raise it. You cut the within-cohort
censoring sweep from this deck, so this concession is your only defence --
deliver it deliberately, not defensively.
:::

# The decomposition {.section-break}

## It is an identity, not an estimator {.smaller .big-math}

Any concordance index that averages over comparable pairs

$$C_w = \frac{\sum_{ij} w_{ij}\,c_{ij}}{\sum_{ij} w_{ij}}$$

splits **exactly**:

$$C_w = \frac{W_{ee}\,C_{ee} \;+\; W_{ec}\,C_{ec}}{W_{ee} + W_{ec}}$$

::: {.fragment .highlight-q}
No new estimator. No added assumptions.<br>
**You already computed these. You averaged them away.**
:::

::: {.fragment}
Matches `survival::concordance()` to machine precision, every tie pattern.
:::

## You cannot escape this by... {.smaller}

::: {.fragment .escape-row}
**...using a better estimator.**
The identity holds for *any* pair-weighted C: Harrell, Uno's IPCW,
time-truncated. Uno's C *reweights* the pairs. It still
[pools]{.pool} them.
:::

::: {.fragment .escape-row}
**...using a better model.**
The decomposition consumes three vectors: time, status, risk score. It does
not know what produced the score. Cox, gradient boosting, random survival
forest and a deep net all decompose identically.
:::

::: {.fragment .escape-row}
**...because [pooling]{.pool} is a convention, not a law.**
Nobody chose it deliberately. Subgroup reporting is routine elsewhere in the
very same papers. The ask is a *split* of the metric you already use.
:::

::: {.fragment .highlight-q}
Masking is a property of **the metric and the cohort**, not of the model
class. These five are all Cox only so the comparison is like-for-like.
:::

::: {.notes}
The middle bullet is aimed squarely at this room: a student audience assumes
this is a classical-regression problem and their ML pipeline is exempt.
It is not.
:::

## What this does not do {.smaller}

- Assumes **non-informative censoring**.
- `C_ee` is imprecise when event-event pairs are scarce: pair counts
  describe *composition*, not *precision*. (`lung`: gap CI
  [`r fmt(ladder$gap_lo[ladder$cohort == "lung"], 2)`,
  `r fmt(ladder$gap_hi[ladder$cohort == "lung"], 2)`] on n = 167.)
- **Discrimination only.** Says nothing about calibration.
- Model-based estimators like **Gönen–Heller are out of scope by
  construction**: they integrate over an assumed model rather than counting
  observed pairs, so the event-event partition is undefined for them.

::: {.notes}
Volunteer these. A peer room respects the speaker who names their own limits
before the audience does, and it removes the cheapest questions.
:::

## Check your C-index {.smaller}

::: {.checklist}
::: {.check}
[Report the censoring rate beside the score.]{.do}
[printed in every result header]{.how}
:::

::: {.check}
[Report `C_ee`, not the pooled number alone.]{.do}
[`fit$C_ee`]{.how}
:::

::: {.check}
[Say whether the gap is distinguishable from zero.]{.do}
[`confint(fit)`]{.how}
:::

::: {.check}
[Show it is not an artefact of the estimator.]{.do}
[`weights_uno()`]{.how}
:::

::: {.check}
[Compare models on the same split, not the pooled score.]{.do}
[`compare_decompositions()`]{.how}
:::
:::

::: {.fragment .highlight-q}
All five come off one call: `decompose_cindex(time, status, risk_score)`.
A stable global C under heavy censoring is **not** evidence of a stable model.
:::

::: {.install}
`pak::pak("Lainsm/cindexdecomp")` · [lainsm.github.io/cindexdecomp](https://lainsm.github.io/cindexdecomp)
:::

# Backup {visibility="uncounted"}

## Does Uno's weighting rescue it? {.smaller visibility="uncounted"}

```{r}
#| label: uno-ladder
#| cache: true
set.seed(2026)
uno_row <- function(label, time, status, lp) {
  f <- decompose_cindex(time, status, lp, weights = weights_uno())
  data.frame(
    Cohort = label, `C_ee` = f$C_ee, `C_ec` = f$C_ec,
    `C_global` = f$C_global, Gap = f$gap, check.names = FALSE
  )
}
uno <- rbind(
  uno_row("lung", lung2$time, lung2$status - 1, predict(m_lung, type = "lp")),
  uno_row("colon", colon2$time, colon2$status, predict(m_colon, type = "lp")),
  uno_row("gbsg", gbsg$rfstime, gbsg$status, predict(m_gbsg, type = "lp")),
  uno_row("rotterdam", rotterdam$dtime, rotterdam$death,
          predict(m_rott, type = "lp")),
  uno_row("flchain", flc$futime, flc$death, predict(m_flc, type = "lp"))
)
uno[, 2:5] <- lapply(uno[, 2:5], fmt)
knitr::kable(uno, align = "lrrrr")
```

Same pattern. Reweighting the pairs does not stop them being pooled.

## Weightings {.smaller visibility="uncounted"}

| Constructor | Estimator |
|---|---|
| `weights_harrell()` | Harrell's C |
| `weights_uno()` | Uno's IPCW C |
| `weights_truncated(tau)` | C truncated at `tau` |
| `weights_custom(fn, name)` | any user-supplied rule |

Under tied event times Uno's pairwise estimator diverges from the
counting-process form in `survival` by around `1e-04`, immaterial in
practice and explained in `vignette("cindexdecomp")`.

## The models {.smaller visibility="uncounted"}

| Cohort | Model |
|---|---|
| `lung` | `age + sex + ph.ecog` |
| `colon` | `age + sex + obstruct + nodes + differ + extent + rx` |
| `gbsg` | `age + meno + size + grade + nodes + pgr + er + hormon` |
| `rotterdam` | `+ chemo` |
| `flchain` | `age + sex + kappa + lambda + mgus` |

All Cox, all complete cases, `n_boot = `r N_BOOT``, `set.seed(2026)`.
All five ship with the `survival` package.
